---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# An R workflow for curation of Philippine Atmospheric, Geophysical, and Astronomical Services Administration (PAGASA) datasets

<!-- badges: start -->
[![License for code](https://img.shields.io/badge/license (for code)-GPL3.0-blue.svg)](https://opensource.org/licenses/gpl-3.0.html)
[![License for data](https://img.shields.io/badge/license (for data)-CC0-blue.svg)](https://creativecommons.org/publicdomain/zero/1.0/)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![test targets workflow](https://github.com/panukatan/pagasa_db/actions/workflows/test-targets-workflow.yml/badge.svg)](https://github.com/panukatan/pagasa_db/actions/workflows/test-targets-workflow.yml)
[![test targets climate](https://github.com/panukatan/paglaom/actions/workflows/test-targets-climate.yml/badge.svg)](https://github.com/panukatan/paglaom/actions/workflows/test-targets-climate.yml)
[![test targets cyclones](https://github.com/panukatan/paglaom/actions/workflows/test-targets-cyclones.yml/badge.svg)](https://github.com/panukatan/paglaom/actions/workflows/test-targets-cyclones.yml)
[![test targets dam](https://github.com/panukatan/paglaom/actions/workflows/test-targets-dam.yml/badge.svg)](https://github.com/panukatan/paglaom/actions/workflows/test-targets-dam.yml)
[![deploy targets dam](https://github.com/panukatan/paglaom/actions/workflows/deploy-targets-dam.yml/badge.svg)](https://github.com/panukatan/paglaom/actions/workflows/deploy-targets-dam.yml)
[![deploy targets releases](https://github.com/panukatan/paglaom/actions/workflows/deploy-targets-releases.yml/badge.svg)](https://github.com/panukatan/paglaom/actions/workflows/deploy-targets-releases.yml)
[![DOI](https://zenodo.org/badge/782627894.svg)](https://zenodo.org/doi/10.5281/zenodo.10934146)
<!-- badges: end -->

This repository is a [`docker`](https://www.docker.com/get-started)-containerised, [`{targets}`](https://docs.ropensci.org/targets/)-based, [`{renv}`](https://rstudio.github.io/renv/articles/renv.html)-enabled [`R`](https://cran.r-project.org/) workflow for the retrieval, processing, and curation of various [Philippine Atmospheric, Geophysical, and Astronomical Services Administration (PAGASA)](https://www.pagasa.dost.gov.ph/) publicly available datasets.

## Why `paglaom`?

The word `paglaom` (pronounced as */paɡˈlaʔom/, [pʌɡˈl̪a.ʔɔm]*) is Bisaya (one of up to 187 languages spoken in the Philippines in addition to Filipino, which is the national language, and English, which is the language of instruction in the country) for hope. [PAGASA](https://pagasa.dost.gov.ph), the national meteorological and hydrological services agency of the Philippines, draws its name from the Filipino word *pag-asa* which means hope. The repository name choice, hence, is a play on these words and also a way to showcase the richness and diversity that exists in the Pilippines.

The `paglaom` project aims to maintain a database of curated datasets on varios atmospheric, geophysical, and astronomical phenomena that are made publicly available by PAGASA on their website. These datasets tend to be summaries of the multitude of data that PAGASA collects on a high frequency basis. They also tend to be in formats that are not machine-readable (e.g., PDF, PNG, HTML formats) meant for reporting to the Philippine population. PAGASA does provide more granular and expansive datasets through a specific data request process. The `paglaom` project aims to showcase publicly available PAGASA data that can be used for various purposes some of which are:

* for students who need to make a report on topics covered by PAGASA's summarised data for a school assignment or project;

* for individuals who have specific interest in one of the natural phenomena that PAGASA monitors and would like to get raw summarised data in a format that is usable and transferrable into other formats;

* for data visualisation learners and aficionados who want to try on working on data about the various natural phenomena available from PAGASA and create unique and interesting plots and graphics.

The broader and more blue skies vision of the `paglaom` project is to contribute to the increasing interest in science, technology, engineering, and mathematics (STEM) subjects particularly in the Philippines with a collection that showcases topics and data that are homegrown and embedded into the fabric of Philippine life.

Whilst the `paglaom` project by its name and the nature of the data it curates has an inherent Filipino audience, it is hoped that those outside of the Philippines will also find the information within useful in similar contexts described above.

## Repository Structure

The project repository is structured as follows:

```
paglaom
    |-- .github/
    |-- data/
    |-- data-raw/
    |-- outputs/
    |-- R/
    |-- reports
    |-- renv
    |-- renv.lock
    |-- .Rprofile
    |-- packages.R
    |-- _targets_climate.R
    |-- _targets_cyclones.R
    |-- _targets_dam.R
    |-- _targets_heat.R
    |-- _targets_setup.R
    |-- _targets.R
    |-- _targets.yaml
```

* `.github` contains project testing and automated deployment of outputs workflows via continuous integration and continuous deployment (CI/CD) using Github Actions.

* `data/` contains intermediate and final data outputs produced by the workflow.

* `data-raw/` contains raw datasets downloaded from publicly available PAGASA sources that are used in the project. This directory is empty given that the raw datasets from PAGASA are in large file size formats that are not ideal for git versioning hence they are git ignored. This directory is kept here to maintain reproducibility of project directory structure and ensure that the workflow runs as expected when run locally.

* `outputs/` contains compiled reports and figures produced by the workflow.

* `R/` contains functions developed/created specifically for use in this workflow.

* `reports/` contains literate code for R Markdown reports rendered in the workflow.

* `renv/` contains `renv` package specific files and directories used by the package for maintaining R package dependencies within the project. The directory `renv/library`, is a library that contains all packages currently used by the project. This directory, and all files and sub-directories within it, are all generated and managed by the `renv` package. Users should not change/edit these manually.

* `renv.lock` file is the `renv` lockfile which records enough metadata about every package used in this project that it can be re-installed on a new machine. This file is generated by the `renv` package and should not be changed/edited manually.

* `.Rprofile` file is a project R profile generated when initiating `renv` for the first time. This file is run automatically every time R is run within this project, and `renv` uses it to configure the R session to use the `renv` project library.

* `packages.R` file lists out all R package dependencies required by the workflow.

* `_targets*.R` files define the steps in the workflow's data ingest, data processing, data analysis, and reporting pipelines.

* `_targets.yaml` file defines the different targets sub-projects within this project.


## Reproducibility

### R package dependencies

This project was built using `R 4.4.1`. This project uses the `renv` framework to record R package dependencies and versions. Packages and versions used are recorded in `renv.lock` and code used to manage dependencies is in `renv/` and other files in the root project directory. After cloning this repository, start an R session in the project's working directory and then run 

```R
renv::restore()
```

to install all R package dependencies.

### Running the workflow

Currently, the project has workflows that curate the following datasets:

1. Tropical cyclones data for various cyclones entering the Philippine area of responsibility since 2017;

2. Daily heat index data from various data collection points in the Philippines;

3. Climatological extremes and normals data over time; and,

4. Daily dam water level data.

The following diagram illustrates these workflows

```{r overall_workflow, echo = FALSE, message = FALSE, results = 'asis'}
cat(
  "```mermaid",
  targets::tar_mermaid(
    targets_only = TRUE, 
    names = dplyr::starts_with(c("climate", "cyclone", "dam", "heat")),
    color = FALSE, legend = FALSE
  ),
  "```",
  sep = "\n"
)
```

To run any of these workflows, run the following command on the R console:

```R
targets::tar_make(dplyr::starts_with("PREFIX"))
```

replacing `"PREFIX"` with the keyword for the type of data. For example, to run the cyclones workflow from R console:

```R
targets::tar_make(dplyr::starts_with("cyclone"))
```

or from the command line/terminal as follows:

```Bash
Rscript -e "targets::tar_make(dplyr::starts_with('cyclone'))"
```

Running specific components of a workflow involves specifying a target name or target names of the components you want to run. You should be able to run a full workflow path by just specifying the name of the last target in the workflow sequence. For example, the following will run the entire cyclones data workflow (as an alternative to what is shown above):

```R
targets::tar_make(cyclones_peak_data_csv)
```

or from the command line/terminal as follows:

```Bash
Rscript -e "targets::tar_make(cyclones_peak_data_csv)"
```

The target `cyclones_peak_data_csv` is the last target of the cyclones data workflow. Hence, to be able to produce the `cyclones_peak_data_csv` target requires running this series of linked targets.

If you would like to run a set of interrelated but not fully linked targets, you will need to specify more than one target name. For this, you can use `tidyselect` approaches to name targets to be run. For example:

```R
targets::tar_make(dplyr::starts_with(c("cyclone", "dam"))
```

will run all targets in the cyclones and dam levels data workflow.

The project also has a workflow for weekly GitHub release of the various raw datasets.

```{r release_workflow, echo = FALSE, message = FALSE, results = 'asis'}
cat(
  "```mermaid",
  targets::tar_mermaid(
    targets_only = TRUE, names = dplyr::starts_with("paglaom_weekly"),
    color = FALSE, legend = FALSE
  ),
  "```",
  sep = "\n"
)
```


## Author

* [Ernest Guevarra](https://github.com/ernestguevarra)


## Licenses

All code created through this project (found in this repository) is released under a [GPL-3.0 license](https://opensource.org/licenses/gpl-3.0.html) license.

Data provided through this project are released under a [CC0](https://creativecommons.org/publicdomain/zero/1.0/) license.

